---
title: "Independent contrasts of sex-specific body mass measurements in spiders"
author: "Jeremy Miller & Rutger Vos"
date: "28/11/2022"
output:
  html_document:
    df_print: paged
---

# Introduction

When analysing trait measurements (such as length) across sets of species we 
have to keep in mind that biological species are not independent entities. All 
life forms are related to one another; all sit somewhere on the tree of life. 
Closely related species probably resemble each other to some extent merely 
because they share common ancestors from which they inherit many of their 
attributes. This shared history is problematic when doing statistical analyses, 
as many methods assume all observations are independent of one another. One way 
to deal with this is to analyse something different: not the direct 
measurements, but the contrasting amounts of evolutionary change the different 
lineages accumulated after they branched off from one another. To figure out 
what those amounts of independent change are, we must know the branching 
pattern. Hence we need to involve the phylogenetic history in our analysis. In 
this tutorial, we demonstrate how that works in the R programming language, 
which most of you have encountered in at least one previous statistics course. 
All the code, data, and results of this tutorial are available here:
https://github.com/naturalis/sysbio-independent-contrasts

# Loading raw measurements

Here we load the measurement data. This input file is in comma-separated format,
with a header that contains column titles. We load this data from the web and
give some summary statistics:

```{r csv}
length_data <- read.csv("https://raw.githubusercontent.com/naturalis/sysbio-independent-contrasts/main/data/MeasurementsPractical.csv")
row.names(length_data) <- length_data$Species
summary(length_data)
```

It appears we have 111 species. For each of those species we have the length of
Males and Females, as well as the Ratio between sexes. The lower bound for body 
length seems not to differ much between sexes, while on the upper extreme, 
females can become substantially larger in some species. However, we have to 
consider this in the context of their evolutionary history.

# Loading tree data

Let's now import a phylogeny. Phylogenetic data comes in a variety of file 
formats (and there are dialects within formats!). None of these file types are
read by the base install of R: we have to use an additional package. Here we
will use the `ape` package, which has a way of representing tree shapes that
has been adopted by many other packages, so it is kind of an informal standard.

```{r tree}
library(ape) # if you don't have this, do 'install.packages("ape")'
phy <- read.nexus('https://raw.githubusercontent.com/naturalis/sysbio-independent-contrasts/main/data/multi_spiders_musc_Beast_10m.tree')
plot.phylo(phy, show.tip.label = F)
```

The above tree is a rooted, ultrametric tree, i.e. there is a sense of 
evolutionary direction (which nodes came first, which came later) and there is
an estimate of timing such that all extant lineages line up in the present.
Let's think this through:

1. For independent contrasts, does the tree have to be rooted? If so, why?
2. Does the tree have to have branch lengths?
3. Does the tree have to be ultrametric?

# Reconciling measurements and tree

We now have a tree object that internally has a somewhat complex structure
(in Rstudio you can explore this in the 'Environment' tab), which includes a
vector of tip labels. We also have a data frame where each row corresponds with
a species. The data frame is not in the same order as the vector of tip labels.
We have to make sure that R knows what tip corresponds with what row in the data
frame otherwise our analysis will be nonsensical. Hence, we are going to line up 
the two data sources and make sure we have the same entities, with
the same names, in both. An easy way to do this is using the `picante` package,
which can interact with trees read by `ape`:

```{r map}
library(picante)
sorted <- match.phylo.data(phy, length_data)
phy <- sorted$phy
length_data <- sorted$data
```

There were some taxa in the tree not in the data (and vice versa). We have now
pruned the data so that we just have the intersection between the two.

# Plotting data

Now that our data are reconciled, let's plot the measurements for males and 
females separately and see if there is any phylogenetic structure to this:

```{r plot}
library(phytools)
# make named vectors
male_length <- as.numeric(length_data$Male)
female_length <- as.numeric(length_data$Female)
names(male_length) <- length_data$Species
names(female_length) <- length_data$Species

# make two continuous trait mappings
male_map <-contMap(phy, male_length, plot=F)
female_map <-contMap(phy, female_length, plot=F, invert=T)

# set up the plot
par(mfrow=c(1,2))
plot(male_map,lwd=3,ftype="off")
plot(female_map,lwd=3,direction="leftwards",ftype="off")
```

The mirrored plot shows male length on the left and female on the right. Let's 
interpret:

1. Can you see phylogenetic structure to female length? Do closely related
   species resemble each other?
2. If 'large body' is green or blue, how many times did large bodies evolve 
   independently?
    
# Calculating independent contrasts

We are now going to calculate the independent contrasts for male length and 
female length

```{r pic}
male_pic <- pic(male_length, phy)
female_pic <- pic(female_length, phy)
plot(male_pic, female_pic)
```

1. Why is there one fewer contrast than there were species in our (pruned) data set?
2. How can there be negative values? Is there a way to make those positive?
3. Try to formulate what has happened in certain contrasts: "In some cases,
   small changes in *this contrast* corresponded with large changes in
   *that contrast*".
   
# Regression analysis
   
Now let's do a simple regression. Note that we have to force the intercept of
the regression line through the origin (why?), hence the extra term in the 
formula:

```{r regress}
regress <- lm(female_pic ~ male_pic - 1)
summary(regress)
```

We can say with confidence that male length is a highly significant (***) 
predictor of female mass. We can also see that this relationship is positive.
Across the entirety of the data set, in most cases any change in male length 
(evolutionary increase or decrease) is matched by a change in the same direction
- hence positive coefficient - and of approximately the same magnitude.

Let's plot the regression line:

```{r line}
plot(male_pic, female_pic)
abline(lm(female_pic ~ male_pic - 1))
```

It should be reasonably clear that there are only very few points in this 
system where one sex changes greatly relative to the other. For example, there
is a single point where the absolute residual is greater than two (and in fact
is nearly three). In human terms, this means that there is a single point on 
the tree where females became nearly three times the length of males. In other
words, one event whose outcome was then retained in all descendants. Can you
point out on the tree where that event took place?

# Conclusions

- Phylogenetic independent contrasts make us look at the direction and 
  magnitude of change on a tree rather than at raw values.
- The contrasts can be analysed in a regression analysis provided we constrain
  the regression line through the origin.
- The residuals that we thus obtain can point us towards events of noticeable
  magnitude, the results of which might be carried around by many descendants.